---
title: "Alpha Phylogenetic Diversity Indices"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Alpha Phylogenetic Diversity Indices}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
ci <- identical(Sys.getenv("CI"), "true")

run_terra <- !ci
run_leaflet <- !ci
```

<br>
[Omar Daniel Leon-Alvarado](https://leon-alvarado.weebly.com/)¹²; [J. Angel Soto-Centeno](https://www.mormoops.com/)²

¹ Department of Earth and Environmental Science, Rutgers University-Newark, NJ, USA.

² Department of Mammalogy, American Museum of Natural History, New York City, NY, USA.
<br>

---

<b>Phylogenetic Diversity (PD)</b> indices are a set of metrics that quantify the <b>evolutionary distinctiveness</b> of a species or a group of species using the information contained within a phylogenetic tree. These indices serve as <b><i>valuable complements</i></b> to traditional measures of ecological diversity (such as species richness or abundance). They enable researchers to identify areas or lineages that harbor high evolutionary heritage. In a rapidly changing world where many species are on the brink of extinction, such evolutionary unique groups may play critical roles in overcoming major environmental crises and maintaining the planet's ecological equilibrium.

The <b>vilma package</b> provides a suite of phylogenetic diversity indices, each with its own logic and interpretation. Because different indices answer different biological questions, users must choose the most appropriate ones for their study. Here, you will learn how to calculate these indices using real‑world data from the <b>Primates of Madagascar</b>.

---

### 0. The Indices

First, you need to know and understand the basics of each indices available in vilma. At this version (1.0) vilma offers six different PD indices, and the option to calculate species richness. 

<br>


```{r, warning=FALSE, include=FALSE}
library(knitr)
library(kableExtra)

df <- data.frame(
  Index = c(
    "Species richness",
    "Faith’s Phylogenetic Diversity",
    "Mean Pairwise Distance",
    "Mean Nearest Taxon Distance",
    "Phylogenetic Endemism",
    "Relative Phylogenetic Endemism",
    "Rao’s Quadratic Entropy",
    "Net Relatedness Index",
    "Nearest Taxon Index"
  ),
  Definition = c(
    "Number of species recorded in each grid cell. Although it is not a phylogenetic metric, it is calculated as a basic reference for comparison with phylogenetic diversity patterns.",
    "Sum of the branch lengths connecting all species present in a grid cell. Higher values indicate that the cell contains a larger amount of evolutionary history.",
    "Average phylogenetic distance between every possible pair of species in a grid cell. It reflects the overall evolutionary relatedness of the assemblage.",
    "Average phylogenetic distance between each species and its closest relative within the grid cell. It is particularly sensitive to relationships near the tips of the phylogeny.",
    "Measures the amount of phylogenetic diversity restricted to a geographic area. Branch lengths are weighted by the geographic range of the descendant species, giving more importance to range-restricted evolutionary lineages.",
    "Compares observed phylogenetic endemism with the value obtained from a comparison tree in which all branches have equal length. It helps identify areas dominated by unusually long or unusually short range-restricted branches.",
    "Measures phylogenetic diversity by combining the phylogenetic distances among species with their relative abundances. Higher values indicate assemblages composed of abundant and evolutionarily distinct species.",
    "Standardized effect size of MPD relative to a null model. It evaluates whether species in a grid cell are more closely related or more distantly related than expected by chance across the entire phylogeny.",
    "Standardized effect size of MNTD relative to a null model. It evaluates phylogenetic clustering or overdispersion near the tips of the phylogeny."
  ),
  Function = c(
    "points_to_raster",                # no function for species richness
    "faith.pd",
    "mpd.calc",
    "mntd.calc",
    "pe.calc",
    "rpe.calc",
    "rao.calc",
    "nri.calc",
    "nti.calc"
  )
)
```
```{r, echo=FALSE}
kable(df, 
      caption = "Table 1: Phylogenetic diversity indices available in the vilma package",
      align = "l") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE,
                position = "center") %>%
  column_spec(2, width = "10cm")  %>%
  row_spec(0, align = "c")
```

<br>

### 1. Data

Now that you know more about PD indices, it is time to proceed with the analyses. The spatial data consist of available occurrence records for all Primates species from Madagascar, obtained from [<b>GBIF</b>](https://doi.org/10.15468/dl.2ah94z). And for the phylogenetic tree, you will use the tree of Primates species from Madagascar available from [VertLife](https://vertlife.org/).


First, load all needed libraries:

```{r, message=FALSE, warning=FALSE}
library(vilma)
library(ape)
library(phytools)
library(sf)
library(terra)
library(dplyr)
library(ggtree)
library(tmap)
library(ggplot2)
library(patchwork)
library(reshape2)
```

<br>

#### 1.1. GBIF Data

In this exercise, you will perform standard spatial data analyses, which require selecting specific columns and filtering occurrences. First, read the downloaded data (a **tab‑separated file** – note that many people mistakenly expect commas and fail). Then, remove columns lacking coordinate information and rows not identified to the species level. Finally, extract the three columns required by vilma: **Species**, **Longitude**, and **Latitude**.

<br>

```{r, eval=FALSE}

# Read the file 
gbif_occ <- read.delim("../inst/extdata/Primates_Mad_GBIF.csv", na.strings = "")

# Check if you have any NA observation
any(is.na(gbif_occ$species))
any(is.na(gbif_occ$decimalLongitude))
any(is.na(gbif_occ$decimalLatitude))

gbif_occ <- gbif_occ[!is.na(gbif_occ$species), ]
```
```{r, echo=FALSE}
# Read the file 
gbif_occ <- read.delim("../inst/extdata/Primates_Mad_GBIF.csv")
gbif_occ <- gbif_occ[!is.na(gbif_occ$species), ]

```
```{r}

# Exrtract the columns and check thm
prim_occ <- data.frame(Species = gbif_occ$species,
                           Lon = gbif_occ$decimalLongitude,
                           Lat = gbif_occ$decimalLatitude)

head(prim_occ, 5L)

```

<br>

Some occurrence records were not identified to the **species level** and therefore needed to be removed. After filtering, the dataset decreased from 17,489 to 17,278 records, which represents only a minor reduction. Although researchers often remove duplicate occurrences, this step is not mandatory in this case. The indices are calculated based on species presence or absence within each raster cell. Therefore, multiple records of the same species within a cell do not affect the results. However, the dataset should still be checked for other issues that are relatively easy to correct.

One of the most common problems is that occurrence records located near the coastline may appear in the sea because of GPS inaccuracies or spatial-resolution errors. The simplest solution would be to remove these records. However, because they may provide useful information, you will attempt to recover them by moving each point to the nearest location on the mainland.

Thus the first step here is to read the polygon map, in this case the boundaries of Madagascar. Then, you will transform the occurrences into a spatial object with the same projection as the polygon.

<br>

```{r}
# Read the polygon
ma_pol <- read_sf("../inst/extdata/Madagascar.gpkg")

# Transform points into a spatal object
prim_occ <- prim_occ %>% st_as_sf(coords = c("Lon","Lat"), crs = 4326)

# Match the projection between the polygon and the points
prim_occ <- st_transform(prim_occ, st_crs(ma_pol))

```

<br>

Now, using the spatial object and the mainland polygon, you will intersect both layers to identify which occurrence records fall **inside** and **outside** the mainland. Once the records have been classified, plot the results using `tmap` to visually inspect their locations.

<br>
```{r, warning=FALSE, message=FALSE, fig.align='center', fig.width=7, fig.height=5}

# Extract the first column from the st_intersect result. 
# This is a boolean vector (TRUE (inside), FALSE (outside))
occ_intersect <- st_intersects(prim_occ, st_union(ma_pol), sparse = F)[,1]

# Split the data
occ_inside <- prim_occ[occ_intersect,]
occ_outside <- prim_occ[!occ_intersect,]

# Plot the results
tm_shape(ma_pol)+ # Plot the polygon
  tm_polygons()+
tm_shape(occ_inside)+ # Add inside points
  tm_dots()+
tm_shape(occ_outside)+ # Add outside points
  tm_dots(fill="red")
```
<br>

Now, let's fix those outside points. First, you will create a line between each point and the country border using `st_nearest_points`. This line represents the shortest distance between the occurrence and the polygon boundary. Then, using `st_cast`, you will convert each line into two points: the starting point (the original occurrence) and the ending point (the nearest location on the border). Finally, by using a grouping index, you will extract the second point of each line—the endpoint—which corresponds to the corrected coordinate located inside the country boundaries.

<br>

```{r, warning=FALSE, message=FALSE, fig.align='center', fig.width=7, fig.height=5}
# Create lines
nearest_lines <- st_nearest_points(occ_outside, st_union(ma_pol))

# Transform lines into points (start and end)
all_points <- st_cast(nearest_lines, "POINT")

# Create index ofr each point: 1,1,2,2,3,3...
point_groups <- rep(1:nrow(occ_outside), each = 2)

# Extract only the second point (end) and transform it into a spatial object
border_points <- all_points[seq(2, length(all_points), by = 2)] %>%  st_as_sf()

# Plot
tm_shape(ma_pol)+ # Add polygon
  tm_polygons()+
tm_shape(occ_inside)+ # Add inside points
  tm_dots()+
tm_shape(occ_outside)+ # Add outside points
  tm_dots(fill="red")+
tm_shape(border_points)+ # Add fixed points
  tm_dots(fill="blue")

# Return spatial points into a data.frame (inside)
occ_inside <- cbind(st_drop_geometry(occ_inside),st_coordinates(occ_inside))

# Return spatial points into a data.frame (fixed)
border_points <- cbind(st_drop_geometry(occ_outside), st_coordinates(border_points))

# Join inside and fixed points
occ_final <- rbind(occ_inside, border_points)

```
<br>

Now you finish to fix and clean your points. However, this is a basic step, usually it is recommend a more deep and detailed cleaning, but with this your a ready to go for the next step.

<br>

#### 1.2 Phylogeny

For phylogenetic diversity indices, you need a **phylogeny** that represents the evolutionary history of the species and allows you to quantify their evolutionary distinctiveness. Therefore, the phylogeny must meet some minimum requirements. The first and most important requirement is that the phylogeny must have **branch lengths**, whether they are calibrated in time or represent genetic/change distances. The type of branch length you use depends on your research question, but remember that each type represents a different biological meaning and may lead to different interpretations. Some indices, such as Faith’s PD, require a **rooted phylogeny**. Although other indices do not necessarily require the tree to be rooted, I strongly recommend always using a rooted phylogeny.

In this step, you will read the tree file. In this case, the file contains a set of 100 trees for the primates of Madagascar, so you will need to calculate a final **consensus tree**. To estimate the branch lengths of the consensus tree, you will use the function `consensus.edges` from the `phytools` package, although `nnls.tree` from the `phangorn` package can also be used. Additionally, you will modify the species names and save them in the phylogeny as `tip.label`. Specifically, you will replace underscores with single spaces so that the species names follow the same format used in the occurrence dataset.


<br>

```{r, message=FALSE, warning=FALSE, fig.align='center', fig.width=7, fig.height=14}
# Read tree
prim_trees <- read.nexus("../inst/extdata/Prim_Madgascar_Tree.nex")

# Majority rule, you can change p to a higher value for a stricter tree
cons_topology <- consensus(prim_trees, p = 0.5)

# Calcualte branch length consensus
prim_tree <- consensus.edges(prim_trees, consensus.tree = cons_topology)

# Change underscore to space in the species' name
prim_tree$tip.label <- gsub("_"," ", prim_tree$tip.label)

# Root the tree, in this case mid point
prim_tree <- midpoint.root(prim_tree)

p <- ggtree(prim_tree) + 
     geom_treescale(fontsize = 1) +           # Add a scale bar
     theme_tree2()+                # Add a subtle background and axis
     geom_tiplab(size = 3)

# Display the plot
print(p)
```
<br>

### 2. vilma.dist object

Now that the occurrence data and phylogeny are ready to use, you will start with the first step: creating a `vilma.dist` object using the function `points_to_raster`. The vilma package is raster-oriented, so most results are returned as raster layers. In this step, you will generate the raster for your study area and define the resolution, or pixel size.

For the pixel resolution, the function only accepts values in **decimal degrees**. For example, if you plan to use the results together with WorldClim or CHELSA variables at **30-arc-second** resolution, you should use **0.008333 degrees**. The same logic applies to the other common resolutions: 2.5 arc-minutes = 0.041667, 5 arc-minutes = 0.083333, and 10 arc-minutes = 0.166667. Once you create the vilma.dist object, you can check the results using the function **plot** or **view.vilma** for an interactive view.

<br>

```{r, fig.align='center', fig.dpi=230, eval=run_leaflet}
library(vilma)

# Create the vilma.dist object, for more information try ?points_to_raster
prim_dist <- points_to_raster(points = occ_final, res = 0.8)

# See the results
view.vilma(prim_dist)

# or use plot(prim_dist)
```

<br>

### 3. PD indices

Now, you can calculate all alpha phylogenetic diversity indices available in vilma. Each alpha-PD function creates a `vilma.pd` object, which contains the raster results, including the PD values and species richness values, a table with the PD and richness values for each raster cell, and additional information such as the algorithm or method used. Likewise, when using the print function, each index will display a complete summary of the results, and the function view.vilma to display the maps in a interactive view.


#### 3.1. Faith's PD index

Faith’s Phylogenetic Diversity, usually called Faith’s PD, measures the **total amount** of evolutionary history represented by the species present in a site, community, or raster cell. This is most famous and common index used, and it requires a rooted phylogeny.

In simple terms, Faith’s PD is calculated by adding the branch lengths of the phylogenetic tree that connect all species found in a given cell (**minimum spanning tree**). If a cell contains species that are distantly related, the total branch length will usually be higher. If a cell contains species that are closely related, the total branch length will usually be lower. For example, imagine two raster cells each contain five species. One cell has five species from the same **closely related** genus, while the other has five species from **different** evolutionary lineages. Even though both cells have the same richness, the second cell will likely have a higher Faith’s PD because it represents more evolutionary history.

In vilma, Faith’s PD is calculated for each raster cell using the species present in that cell and the branch lengths of the phylogeny. The result is a raster layer where each pixel contains the Faith’s PD value for that location. A higher value means that the cell contains more accumulated evolutionary history.

Because Faith’s PD is based on the minimum spanning tree connecting the species present in a site or raster cell, it generally requires at least **two species** per cell to calculate the index. Therefore, cells containing only a single species require special treatment. For these cases, the function `faith.pd` provides three options: 1) `exclude`, which removes single-species cells from the analysis; 2) `root`, which assigns the branch length from the species tip to the root of the tree; and 3) `node`, which assigns the branch length from the species tip to its most recent common ancestor with the rest of the tree. The choice among these options depends on the **research question** and the structure of the data. For example, when using large phylogenies, the `root` option may overestimate the contribution of a single species because it includes the full path from the tip to the root.

When the index is calculated, check the message generated by the `faith.pd` function. The function will compare the species names in the `vilma.dist` object with the tip labels in the **phylogeny**. It will then select only the **species shared** by both datasets and return summary information about the matching process. This step is important because phylogenetic diversity indices can only be calculated for species that are present in both the occurrence data and the phylogeny.

<br>
```{r, fig.align='center', fig.dpi=230, eval=run_leaflet}
# Calcualte Faith's PD using the root
prim_pd <- faith.pd(tree = prim_tree, dist = prim_dist, method = "root")

# If you want other options, run these lines
# prim_pd <- faith.pd(tree = prim_tree, dist = prim_dist, method = "node")
# prim_pd <- faith.pd(tree = prim_tree, dist = prim_dist, method = "exclude")

# Check the summary results
print(prim_pd)

# Check the interactive map

view.vilma(prim_pd)
```
<br>

One common issue with Faith’s PD is that this index is often strongly **correlated** with species richness. In general, cells with higher species richness tend to have higher PD values because more species usually add more branches to the minimum spanning tree. You can use the table stored in the resulting `vilma.pd` object to evaluate how strongly PD and species richness are correlated in your dataset.

<br>
```{r, warning=F, fig.align='center', fig.dpi=230}

# Check pearson correlation
cor.test(prim_pd$pd.table$SR, prim_pd$pd.table$PD, method = "pearson")

# Check spearman correlation
cor.test(prim_pd$pd.table$SR, prim_pd$pd.table$PD, method = "spearman")

# Plot the data
ggplot(data = prim_pd$pd.table)+
  geom_point(aes(x = PD, y = SR))+
  geom_smooth(aes(x = PD, y = SR))+
  theme_bw()

```
<br>

As you can see, Faith’s PD is highly correlated with species richness. This means that the index may be capturing mainly the number of species in each cell rather than differences in evolutionary distinctiveness among assemblages. A good alternative is to use indices based on **phylogenetic distance matrices**, such as **Mean Pairwise Distance (MPD)** and **Mean Nearest Taxon Distance (MNTD)**. These indices are less directly dependent on species richness because they summarize the average phylogenetic distance among species, rather than the total amount of branch length accumulated in each cell.

<br>

#### 3.2. Mean Pairwise Distance (MPD) index.

Mean Pairwise Distance (MPD) is an alpha phylogenetic diversity index that measures the **average phylogenetic distance** among all pairs of species occurring in a site, community, or raster cell. Instead of summing the total branch length, as in Faith’s PD, MPD uses a **phylogenetic distance matrix** and calculates the mean distance between every possible pair of species present in the assemblage.

The logic of MPD is to evaluate how evolutionarily related or distinct the species in a cell are from each other. High MPD values indicate that the species present in a cell are, on average, **distantly related**. This means that the assemblage contains species from different evolutionary lineages. Low MPD values indicate that the species are, on average, **closely related**, suggesting that the assemblage is composed of species from more similar or closely related lineages.

In `vilma`, the function `mpd.calc` calculates the Mean Pairwise Distance (MPD) index. Similar to Faith’s PD, this index requires at least two species per cell, so the function implements the same three options for single-species cells: `exclude`, `root`, and `node`. In addition, MPD can incorporate abundance information when available. However, in this example, abundance will not be used, so the abundance argument will be set to `FALSE`.


```{r, fig.align='center', fig.dpi=230, eval=run_leaflet}
# Calculate the index using root method
prim_mpd <- mpd.calc(tree = prim_tree, dist = prim_dist, method = "root", abundance = FALSE)

# Other options
# prim_mpd <- mpd.calc(tree = prim_tree, dist = prim_dist, method = "exclude", abundance = FALSE)
# prim_mpd <- mpd.calc(tree = prim_tree, dist = prim_dist, method = "node", abundance = FALSE)

# Check the summary results
print(prim_mpd)

# Check the interactive map
view.vilma(prim_mpd)
```
<br>

#### 3.3. Mean Nearest Taxon Distance (MNTD)

Mean Nearest Taxon Distance (MNTD) is an alpha phylogenetic diversity index that measures the **average phylogenetic distance** between each species in a site, community, or raster cell and its closest relative within that same assemblage. Unlike MPD, which considers all possible pairs of species, MNTD focuses only on the **nearest phylogenetic neighbor of each species**.

The logic of MNTD is to evaluate how closely related the species in a cell are near the tips of the phylogeny. High MNTD values indicate that each species is, on average, **relatively distant** from its closest relative in the assemblage. This suggests that the cell contains species that are not closely related at the terminal branches of the tree. Low MNTD values indicate that species tend to **have close relatives** within the same cell, suggesting stronger phylogenetic similarity or clustering near the tips of the phylogeny.

MNTD is especially useful for detecting **fine-scale phylogenetic structure**. While MPD summarizes the overall evolutionary distance among all species in a cell, MNTD is more sensitive to recent evolutionary relationships because it focuses on the closest relatives. For this reason, MNTD can help identify whether assemblages are composed of closely related species from the same recent lineages or species that are more evolutionary isolated from their nearest relatives.

In vilma, the function `mntd.calc` calculates the Mean Nearest Taxon Distance (MNTD) index. This function uses the same main parameters as `mpd.calc`, including the options for handling single-species cells: `exclude`, `root`, and  `node`. As with MPD, MNTD can also incorporate abundance information when available. However, in this example, abundance will not be used, so the abundance argument will be set to FALSE.

<br>
```{r, fig.align='center', fig.dpi=230, eval=run_leaflet}
# Calculate the index using root method
prim_mntd <- mntd.calc(tree = prim_tree, dist = prim_dist, method = "root", abundance = FALSE)

# Other options
# prim_mntd <- mntd.calc(tree = prim_tree, dist = prim_dist, method = "exclude", abundance = FALSE)
# prim_mntd <- mntd.calc(tree = prim_tree, dist = prim_dist, method = "node", abundance = FALSE)

# Check the summary results
print(prim_mntd)

# Check the interactive map
view.vilma(prim_mntd)
```
<br>

Now, you can visually compare how these indices differ from each other and how each one is **correlated** with species richness (SR). Of course, every index has its own logic, objective, and **biological interpretation**. However, it is always useful to evaluate the influence of species richness on each index, especially because some phylogenetic diversity metrics may be strongly driven by the number of species present in each cell.

<br>
```{r,message=FALSE, warning=FALSE, fig.align='center', fig.width=7}
# Plot the PD data
pd <- ggplot(data = prim_pd$pd.table)+
  geom_point(aes(x = PD, y = SR))+
  geom_smooth(aes(x = PD, y = SR))+
  theme_bw()+
  ggtitle("Faith's PD")

# Plot the MPD data
mpd <- ggplot(data = prim_mpd$pd.table)+
  geom_point(aes(x = PD, y = SR))+
  geom_smooth(aes(x = PD, y = SR))+
  theme_bw()+
  ggtitle("MPD")

# Plot the MNTD data
mntd <- ggplot(data = prim_mntd$pd.table)+
  geom_point(aes(x = PD, y = SR))+
  geom_smooth(aes(x = PD, y = SR))+
  theme_bw()+
  ggtitle("MNTD")

# Multiple using patchwork library
pd+mpd+mntd

```
<br>

#### 3.4. Phylogenetic Endemism (PE) index

Phylogenetic Endemism (PE), is an alpha phylogenetic diversity index that measures how much evolutionary history is **geographically restricted** to a particular site, community, or raster cell. Unlike Faith’s PD, which sums the total branch length represented by the species in a cell, PE **weights** each branch by its **geographic range**. Therefore, branches that occur in fewer cells contribute more strongly to the index than branches that are widely distributed.

The logic of PE is to identify areas that contain unique or range-restricted evolutionary history. A cell will have high PE when it contains species, or deeper evolutionary lineages, that occur in only a **small part of the study area**. In contrast, a cell may have high species richness or high Faith’s PD but low PE if most of its species belong to widespread lineages. Therefore, PE is especially useful for detecting areas of conservation importance because it highlights places that represent **geographically rare evolutionary diversity**.

In practice, PE combines phylogenetic information and spatial distribution information. Each branch of the phylogeny is evaluated according to the distribution of the species descending from that branch. Branches with restricted geographic ranges receive **higher weight**, while branches shared by many widespread species receive **lower weight**. As a result, PE gives more importance to cells containing endemic or narrowly distributed lineages.

In vilma, the function `pe.calc` calculates the Phylogenetic Endemism (PE) index. The function also includes the argument `RPE`, a logical option that determines whether **Relative Phylogenetic Endemism** should also be calculated. If RPE = TRUE, which is the default, the function computes Relative PE as **RPE = PE / PD**.

Because PE depends on PD, the function also includes the methods `exclude`, `root`, and `node` to handle cells with a single species. These options determine how single-species cells are treated during the calculation. Relative PE expresses the **proportion of phylogenetic diversity** in each cell that is geographically restricted. The resulting raster identifies cells that concentrate range-restricted evolutionary history. High PE values indicate areas where evolutionary diversity is both **phylogenetically meaningful** and **geographically restricted**.

<br>
```{r, fig.align='center', fig.dpi=230, eval=run_leaflet}
# Calculate the index using root method
prim_pe <- pe.calc(tree = prim_tree, dist = prim_dist, RPE = TRUE, faith.method = "root")

# Other options
# prim_pe <- pe.calc(tree = prim_tree, dist = prim_dist, RPE = TRUE, faith.method = "exclude")
# prim_pe <- pe.calc(tree = prim_tree, dist = prim_dist, RPE = TRUE, faith.method = "node")

# Check the summary results
print(prim_pe)

# Check the interactive map
view.vilma(prim_pe)
```
<br>

#### 3.5. Rao’s Quadratic Entropy (RaoQ)

Rao’s Quadratic Entropy, also known as **RaoQ**, is an alpha phylogenetic diversity index that measures the **expected phylogenetic dissimilarity** between two individuals randomly selected from the same site, community, or raster cell. In the context of phylogenetic diversity, RaoQ uses the phylogenetic distance among species to evaluate how **evolutionarily different** the species within a cell are from one another.

The logic of RaoQ is based on both phylogenetic distance and, when available, species abundance. If two randomly selected individuals from a cell are likely to belong to distantly related species, the cell will have a **high** RaoQ value. In contrast, if the individuals are likely to belong to the same species or to closely related species, the cell will have a **low** RaoQ value. Therefore, RaoQ describes within-community phylogenetic dispersion, giving higher values to assemblages composed of evolutionarily distinct lineages.

In vilma, RaoQ is calculated with the function `rao.calc`, using a rooted phylogenetic tree. The function also includes the argument `scale01`. If `scale01 = TRUE`, which is the **default**, the phylogenetic distance matrix is divided by its maximum value so that distances range from 0 to 1. Under this scaling, RaoQ values also range approximately from 0 to 1, which makes results easier to interpret and compare. Larger RaoQ values indicate greater within-cell phylogenetic dispersion, while lower values indicate that the assemblage is composed of more closely related species or dominated by few closely related lineages. The function includes the argument `abundance`. If `abundance = FALSE`, which is the **default**, the calculation uses presence–absence data, meaning that all species present in a cell are weighted equally. If `abundance = TRUE`, the function uses the number of records per species within each cell to estimate relative abundances.

<br>
```{r, fig.align='center', fig.dpi=230, eval=run_leaflet}
# Calculate the index using root method
prim_rao <- rao.calc(tree = prim_tree, dist = prim_dist, abundance = FALSE, scale01 = TRUE)

# Other options
# prim_rao <- prim_rao <- rao.calc(tree = prim_tree, dist = prim_dist, abundance = TRUE, scale01 = TRUE)
# prim_rao <- prim_rao <- rao.calc(tree = prim_tree, dist = prim_dist, abundance = FALSE, scale01 = FALSE)

# Check the summary results
print(prim_rao)

# Check the interactive map
view.vilma(prim_rao)
```
<br>

#### 3.6. Net Relatedness Index (NRI) and Nearest Taxon Index (NTI)

Net Relatedness Index (NRI) and Nearest Taxon Index (NTI) are standardized alpha phylogenetic diversity indices used to evaluate whether the species within each site, community, or raster cell are more closely related or more distantly related than **expected under a null model**. Unlike raw indices such as **MPD** and **MNTD**, these indices compare the observed phylogenetic structure of each cell against randomized expectations.

**NRI** is based on the standardized effect size of **Mean Pairwise Distance (MPD)**. Therefore, it evaluates the overall phylogenetic structure of the assemblage by considering the average phylogenetic distance among all pairs of species in a cell. **Positive** NRI values indicate phylogenetic clustering, meaning that species in the cell are more closely related than expected by chance. **Negative** NRI values indicate phylogenetic overdispersion, meaning that species are more distantly related than expected under the null model. Because NRI is based on MPD, it captures deeper or broader phylogenetic structure across the tree.

**NTI** is based on the standardized effect size of **Mean Nearest Taxon Distance (MNTD)**. Therefore, it focuses on the phylogenetic distance between each species and its closest relative within the cell. **Positive** NTI values indicate clustering near the tips of the phylogeny, meaning that species tend to co-occur with close relatives more often than expected by chance. **Negative** NTI values indicate overdispersion near the tips, meaning that species tend to be more distantly related from their nearest relatives than expected under the null model. Because NTI is based on MNTD, it is more sensitive to recent evolutionary relationships and terminal branches.

In vilma, the functions `nri.calc` and `nti.calc` calculate NRI and NTI, respectively. The function `nri.calc` uses **MPD** as the observed metric, while `nti.calc` uses **MNTD**. Both functions require a rooted phylogenetic tree with branch lengths and a `vilma.dist` object containing the species distributions. They also implement the same options for handling single-species cells: `root`, `node`, and `exclude`. Additionally, both functions can incorporate abundance information, although abundance is set to FALSE by default.

Both **NRI** and **NTI** require null-model simulations. In vilma, the number of randomizations is controlled by the `iterations` argument, with a default of 999. The null model is defined using the sampling argument, which can be set to `taxa.label`, `range`, `neighbor`, or `regional`. These options allow the user to test phylogenetic structure under different randomization assumptions. For example, `taxa.label` randomizes species labels on the phylogeny, while regional builds communities by drawing species from a regional species pool.

The logic of these indices is that raw **MPD** or **MNTD** values alone do not indicate whether an assemblage is unusually clustered or overdispersed. A cell may have low MPD or MNTD simply because the available species pool is composed of closely related species. By comparing the observed value against a null distribution, NRI and NTI identify whether the **observed** phylogenetic structure is **stronger** or **weaker** than expected by chance. Thus, NRI and NTI are especially useful for detecting non-random phylogenetic patterns in community composition.

<br>
```{r, fig.align='center', fig.dpi=230, message=FALSE, warning=FALSE, eval=FALSE}
# Calculate the NTI index using root method
prim_nti <- nti.calc(tree = prim_tree, dist = prim_dist, mntd.method = "root", abundance = FALSE, iterations = 999, sampling = "taxa.label")

# Check the NTI summary results
print(prim_nti)
```

```{r, echo=FALSE, fig.align='center', fig.dpi=230, message=FALSE, warning=FALSE}
# Calculate the NTI index using root method
invisible(
  capture.output(
    prim_nti <- suppressMessages(nti.calc(tree = prim_tree, dist = prim_dist, mntd.method = "root", abundance = FALSE, iterations = 10, sampling = "taxa.label"))
  )
)

# Check the NTI summary results
print(prim_nti)
```

```{r, eval=FALSE}
# Check the NTI interactive map
view.vilma(prim_nti)
```

```{r, echo=FALSE, fig.align='center', fig.dpi=230, message=FALSE, warning=FALSE, eval=run_leaflet}
# Check the NTI interactive map
view.vilma(prim_nti)
```

```{r, fig.align='center', fig.dpi=230, message=FALSE, warning=FALSE, eval=FALSE}
# Calculate the NRI index using root method
prim_nri <- nri.calc(tree = prim_tree, dist = prim_dist, mpd.method = "root", abundance = FALSE, iterations = 999, sampling = "taxa.label")

# Check the NRI summary results
print(prim_nri)
```

```{r, echo=FALSE, fig.align='center', fig.dpi=230, message=FALSE, warning=FALSE}
invisible(
  capture.output( 
    prim_nri <- suppressMessages(nri.calc(tree = prim_tree, dist = prim_dist, mpd.method = "root", abundance = FALSE, iterations = 10, sampling = "taxa.label"))
    )
)

# Check the NRI summary results
print(prim_nri)
```

```{r, eval=FALSE}
# Check the NRI interactive map
view.vilma(prim_nri)

```

```{r, echo=FALSE, fig.align='center', fig.dpi=230, message=FALSE, warning=FALSE, eval=run_leaflet}
# Check the NRI interactive map
view.vilma(prim_nri)
```
<br>

---

Now, you have calculated all the alpha phylogenetic diversity indices currently available in `vilma`. The number of available indices may increase over time, but you have already seen that the functions follow a similar and straightforward structure. The most important point is not only to know how to run each function, but also to understand the logic behind each index, when each index should be used, and what type of ecological or evolutionary question each one can help answer.

To better understand the behavior of the indices, you will perform a simple but useful analysis to evaluate how similar the results are among them. In this case, you will create a heatmap showing the pairwise correlations among all alpha phylogenetic diversity indices calculated for the primates of Madagascar. This analysis will help identify which indices produce similar spatial patterns and which ones capture different aspects of phylogenetic diversity.

<br>
```{r, message=FALSE, warning=FALSE, fig.align='center', fig.width=7, fig.height=7}

# Create the table with the indices values
index_values <- data.frame(SR = prim_pd$pd.table$SR,
                           PD = prim_pd$pd.table$PD,
                           MPD = prim_mpd$pd.table$PD,
                           MNTD = prim_mntd$pd.table$PD,
                           PE = prim_pe$pd.table$PD,
                           RPE = prim_pe$pd.table$RPE,
                           RAOQ = prim_rao$pd.table$PD,
                           NTI = prim_nti$pd.table$PD,
                           NRI = prim_nri$pd.table$PD)

# Create a pairwise correlation matrix
cor_matrix <- cor(index_values, use = "pairwise.complete.obs")

# Melt the matrix for an easy plotting
cor_melt <- melt(cor_matrix)

# Create the plot
ggplot(cor_melt, aes(x = Var2, y = Var1, fill = value)) +
  geom_tile() +
  geom_text(aes(label = round(value, 2)), size = 4) + # Add correlation numbers
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", 
                       midpoint = 0, limit = c(-1, 1)) +
  labs(x = "", y = "", title = "Correlation Heatmap of PD Indices") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
<br>

The correlation matrix reveals three **main patterns**. First, SR, PD, PE, and RPE form a strongly positive block (r ≥ 0.66–0.97), indicating rich sites harbor deep evolutionary history and endemism. Yet PE and RPE are nearly identical (0.97), making RPE empirically redundant. Second, SR strongly correlates with PD (0.91) but negatively with MNTD (-0.57) while showing no link to MPD (0.10), suggesting habitat filtering adds closely related species without altering overall phylogenetic breadth. Third, standardized effect sizes (NTI/NRI) are decoupled from SR (r ≈ 0.06–0.23) but negatively tied to MPD/MNTD, proving phylogenetic structure depends on species identity rather than richness. In short, your correlation matrix has already told you that **richness**, **structure**, and **evenness** are independent stories in your system. 

With this kind of analysis, you can proceed with a more specific analysis, and it will help you decide which indices to use.

---

### 4. Saving results.

Once you have selected the index or indices that you want to keep and use for further analyses, `vilma` allows you to save the alpha phylogenetic diversity results using the function `write.vilma.pd`. This function requires a `vilma.pd` object, a `file` name, which can also include the output directory path, and the `raster.format` argument. The available raster format options are `tif`, `grd`, and 
`.

The function generates different output files. First, it creates a **CSV file** containing the values associated with each raster cell, including species richness, phylogenetic diversity values, and any other index-specific results. Second, it saves the **raster layers** available in the `vilma.pd` object using the raster format specified by the user. Third, it generates a **log file** containing the summary information of the `vilma.pd` object, equivalent to the information displayed when using the print function. Let see this example saving the results of MPD.

<br>
```{r, eval=FALSE}

# Saving the results as
write.vilma.pd(vilma.pd = prim_mpd, file = "results/Prim_Madagascar_MPD", raster.format = "tif", overwrite = TRUE)

```
<br>

The output folder will looks like:

<br>
```{}
results/
  ├── Prim_Madagascar_MPD.csv
  ├── Prim_Madagascar_MPD_pd_raster.tif
  └── Prim_Madagascar_MPD_log.txt
```
<br>


### 5. Conclusion

In this tutorial, you learned how to calculate alpha phylogenetic diversity indices in `vilma` using occurrence data, a phylogenetic tree, and a `vilma.dist` object. You explored several indices, including Faith’s PD, MPD, MNTD, PE, RPE, RaoQ, NRI, and NTI, each representing a different dimension of evolutionary diversity within raster cells.

You also learned how to compare indices and evaluate their relationship with species richness. This step is important because some indices may be strongly influenced by the number of species in each cell, while others capture phylogenetic structure, endemism, or evolutionary distinctiveness more independently.

Finally, you learned how to save the results using `write.vilma.pd`, including raster layers, cell-level tables, and summary log files. Overall, vilma provides a flexible raster-based framework to calculate, visualize, compare, and export alpha phylogenetic diversity patterns across geographic space.
