---
title: "Beta Phylogenetic Diversity Indices"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Beta Phylogenetic Diversity Indices}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)

# By default, do not run optional spatial maps or heavy calculations.
# This keeps R CMD check stable across machines.
run_terra <- identical(Sys.getenv("VILMA_RUN_MAPS"), "true")
run_leaflet <- identical(Sys.getenv("VILMA_RUN_MAPS"), "true")
run_heavy <- identical(Sys.getenv("VILMA_RUN_HEAVY"), "true")
```

<br>
[Omar Daniel Leon-Alvarado](https://leon-alvarado.weebly.com/)¹²; [J.Angel Soto-Centeno](https://www.mormoops.com/)²

¹ Department of Earth and Environmental Science, Rutgers University-Newark, NJ, USA.

² Department of Mammalogy, American Museum of Natural History, New York City, NY, USA.
<br>

---

**Beta diversity** measures the variation in species composition among sites, communities, or raster cells. While alpha diversity describes diversity within a single site, beta diversity focuses on how different sites are from each other. In simple terms, beta diversity helps answer whether two cells contain similar or different sets of species.

**Beta phylogenetic diversity** extends this idea by incorporating the evolutionary relationships among species. Instead of comparing sites only based on shared or unique species, beta phylogenetic diversity compares how much evolutionary history is shared or different among communities. Therefore, two cells may have no species in common but still show low phylogenetic dissimilarity if their species are closely related. Conversely, two cells may be highly distinct if they contain species from distant evolutionary lineages. This makes beta phylogenetic diversity useful for evaluating spatial turnover in evolutionary history and identifying areas that contribute unique phylogenetic composition across the landscape.

---

### 0. The Indices

First, you need to know and understand the basics of each indices available in `vilma`. At this version (1.0) vilma offers six different Beta-PD indices, and the option to calculate species richness. 

<br>

```{r beta-table-setup, warning=FALSE, include=FALSE}
library(kableExtra)

df_beta <- data.frame(
  Index = c(
    "Beta Mean Nearest Taxon Distance",
    "Beta Mean Pairwise Distance",
    "Phylogenetic Beta Diversity",
    "PhyloSor",
    "UniFrac",
    "Rao’s Q Beta Diversity"
  ),
  Definition = c(
    "Measures beta phylogenetic diversity using the nearest phylogenetic-neighbor distances among communities or raster cells. It is based on MNTD and is sensitive to differences near the tips of the phylogeny.",
    "Measures beta phylogenetic diversity using the mean pairwise phylogenetic distances among communities or raster cells. It is based on MPD and captures broader phylogenetic differences across the tree.",
    "Estimates phylogenetic beta diversity by comparing the amount of shared and unique evolutionary history among communities. It can be calculated using weighted or unweighted approaches.",
    "Measures phylogenetic similarity or dissimilarity among communities based on shared branch lengths. It is the phylogenetic equivalent of Sørensen-type similarity and is useful for evaluating shared evolutionary history among cells.",
    "Measures phylogenetic dissimilarity among communities based on the fraction of branch length that is unique to each community. It can be calculated using unweighted or weighted approaches, depending on whether abundance information is included.",
    "Measures beta diversity using Rao’s quadratic entropy framework. It evaluates phylogenetic differences among communities based on pairwise phylogenetic distances and, optionally, abundance information."
  ),
  Function = c(
    "beta.mntd",
    "beta.mpd",
    "phylo.beta",
    "phylosor.calc",
    "unifrac.calc",
    "rao.beta"
  )
)
```

```{r beta-functions-table, echo=FALSE}
kable(
  df_beta,
  caption = "Table 2: Beta phylogenetic diversity indices available in the vilma package",
  align = "l"
) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE,
    position = "center"
  ) %>%
  column_spec(2, width = "10cm") %>%
  row_spec(0, align = "c")
```
<br>

### 1. Data

Now that you know more about PD indices, it is time to proceed with the analyses. The spatial data consist of available occurrence records for all Primates species from Madagascar, obtained from [<b>GBIF</b>](https://doi.org/10.15468/dl.2ah94z). And for the phylogenetic tree, you will use the tree of Primates species from Madagascar available from [VertLife](https://vertlife.org/).


First, load all needed libraries:

```{r, message=FALSE, warning=FALSE}
library(vilma)
library(ape)
library(phytools)
library(sf)
library(terra)
library(dplyr)
library(ggtree)
library(ggplot2)
library(patchwork)
library(reshape2)
```

<br>

#### 1.1. GBIF Data

In this exercise, you will perform standard spatial data analyses, which require selecting specific columns and filtering occurrences. First, read the downloaded data (a **tab‑separated file** – note that many people mistakenly expect commas and fail). Then, remove columns lacking coordinate information and rows not identified to the species level. Finally, extract the three columns required by vilma: **Species**, **Longitude**, and **Latitude**.

<br>

```{r, eval=FALSE}

# Read the file 
gbif_occ <- read.delim("../inst/extdata/Primates_Mad_GBIF.csv", na.strings = "")

# Check if you have any NA observation
any(is.na(gbif_occ$species))
any(is.na(gbif_occ$decimalLongitude))
any(is.na(gbif_occ$decimalLatitude))

gbif_occ <- gbif_occ[!is.na(gbif_occ$species), ]
```
```{r, echo=FALSE}
# Read the file 
gbif_occ <- read.delim("../inst/extdata/Primates_Mad_GBIF.csv")
gbif_occ <- gbif_occ[!is.na(gbif_occ$species), ]

```
```{r}

# Exrtract the columns and check thm
prim_occ <- data.frame(Species = gbif_occ$species,
                           Lon = gbif_occ$decimalLongitude,
                           Lat = gbif_occ$decimalLatitude)

head(prim_occ, 5L)

```

<br>

Some occurrence records were not identified to the **species level** and therefore needed to be removed. After filtering, the dataset decreased from 17,489 to 17,278 records, which represents only a minor reduction. Although researchers often remove duplicate occurrences, this step is not mandatory in this case. The indices are calculated based on species presence or absence within each raster cell. Therefore, multiple records of the same species within a cell do not affect the results. However, the dataset should still be checked for other issues that are relatively easy to correct.

One of the most common problems is that occurrence records located near the coastline may appear in the sea because of GPS inaccuracies or spatial-resolution errors. The simplest solution would be to remove these records. However, because they may provide useful information, you will attempt to recover them by moving each point to the nearest location on the mainland.

Thus the first step here is to read the polygon map, in this case the boundaries of Madagascar. Then, you will transform the occurrences into a spatial object with the same projection as the polygon.

<br>

```{r}
# Read the polygon
ma_pol <- read_sf("../inst/extdata/Madagascar.gpkg")

# Transform points into a spatal object
prim_occ <- prim_occ %>% st_as_sf(coords = c("Lon","Lat"), crs = 4326)

# Match the projection between the polygon and the points
prim_occ <- st_transform(prim_occ, st_crs(ma_pol))

```

<br>

Now, using the spatial object and the mainland polygon, you will intersect both layers to identify which occurrence records fall **inside** and **outside** the mainland. Once the records have been classified, plot the results using `tmap` to visually inspect their locations.

<br>
```{r, warning=FALSE, message=FALSE}
# Extract the first column from the st_intersect result. 
# This is a boolean vector (TRUE (inside), FALSE (outside))
occ_intersect <- st_intersects(prim_occ, st_union(ma_pol), sparse = F)[,1]

# Split the data
occ_inside <- prim_occ[occ_intersect,]
occ_outside <- prim_occ[!occ_intersect,]
```

```{r, warning=FALSE, message=FALSE, fig.align='center', fig.width=7, fig.height=5, eval=run_terra}
library(tmap)

# Plot the results
tm_shape(ma_pol)+ # Plot the polygon
  tm_polygons()+
tm_shape(occ_inside)+ # Add inside points
  tm_dots()+
tm_shape(occ_outside)+ # Add outside points
  tm_dots(fill="red")
```
<br>

Now, let's fix those outside points. First, you will create a line between each point and the country border using `st_nearest_points`. This line represents the shortest distance between the occurrence and the polygon boundary. Then, using `st_cast`, you will convert each line into two points: the starting point (the original occurrence) and the ending point (the nearest location on the border). Finally, by using a grouping index, you will extract the second point of each line—the endpoint—which corresponds to the corrected coordinate located inside the country boundaries.

<br>

```{r, warning=FALSE, message=FALSE}
library(terra)

# Create lines
nearest_lines <- st_nearest_points(occ_outside, st_union(ma_pol))

# Transform lines into points (start and end)
all_points <- st_cast(nearest_lines, "POINT")

# Create index ofr each point: 1,1,2,2,3,3...
point_groups <- rep(1:nrow(occ_outside), each = 2)

# Extract only the second point (end) and transform it into a spatial object
border_points <- all_points[seq(2, length(all_points), by = 2)] %>%  st_as_sf()
```

```{r, warning=FALSE, message=FALSE, fig.align='center', fig.width=7, fig.height=5, eval=run_terra}
library(terra)

# Plot
tm_shape(ma_pol)+ # Add polygon
  tm_polygons()+
tm_shape(occ_inside)+ # Add inside points
  tm_dots()+
tm_shape(occ_outside)+ # Add outside points
  tm_dots(fill="red")+
tm_shape(border_points)+ # Add fixed points
  tm_dots(fill="blue")
```

```{r, warning=FALSE, message=FALSE}
# Return spatial points into a data.frame (inside)
occ_inside <- cbind(st_drop_geometry(occ_inside),st_coordinates(occ_inside))

# Return spatial points into a data.frame (fixed)
border_points <- cbind(st_drop_geometry(occ_outside), st_coordinates(border_points))

# Join inside and fixed points
occ_final <- rbind(occ_inside, border_points)

```
<br>

Now you finish to fix and clean your points. However, this is a basic step, usually it is recommend a more deep and detailed cleaning, but with this your a ready to go for the next step.

<br>

#### 1.2 Phylogeny

For phylogenetic diversity indices, you need a **phylogeny** that represents the evolutionary history of the species and allows you to quantify their evolutionary distinctiveness. Therefore, the phylogeny must meet some minimum requirements. The first and most important requirement is that the phylogeny must have **branch lengths**, whether they are calibrated in time or represent genetic/change distances. The type of branch length you use depends on your research question, but remember that each type represents a different biological meaning and may lead to different interpretations. Some indices, such as Faith’s PD, require a **rooted phylogeny**. Although other indices do not necessarily require the tree to be rooted, I strongly recommend always using a rooted phylogeny.

In this step, you will read the tree file. In this case, the file contains a set of 100 trees for the primates of Madagascar, so you will need to calculate a final **consensus tree**. To estimate the branch lengths of the consensus tree, you will use the function `consensus.edges` from the `phytools` package, although `nnls.tree` from the `phangorn` package can also be used. Additionally, you will modify the species names and save them in the phylogeny as `tip.label`. Specifically, you will replace underscores with single spaces so that the species names follow the same format used in the occurrence dataset.


<br>

```{r, message=FALSE, warning=FALSE, fig.align='center', fig.width=7, fig.height=14}
# Read tree
prim_trees <- read.nexus("../inst/extdata/Prim_Madgascar_Tree.nex")

# Majority rule, you can change p to a higher value for a stricter tree
cons_topology <- consensus(prim_trees, p = 0.5)

# Calcualte branch length consensus
prim_tree <- consensus.edges(prim_trees, consensus.tree = cons_topology)

# Change underscore to space in the species' name
prim_tree$tip.label <- gsub("_"," ", prim_tree$tip.label)

# Root the tree, in this case mid point
prim_tree <- midpoint.root(prim_tree)

p <- ggtree(prim_tree) + 
     geom_treescale(fontsize = 1) +           # Add a scale bar
     theme_tree2()+                # Add a subtle background and axis
     geom_tiplab(size = 3)

# Display the plot
print(p)
```
<br>

### 2. vilma.dist object

Now that the occurrence data and phylogeny are ready to use, you will start with the first step: creating a `vilma.dist` object using the function `points_to_raster`. The vilma package is raster-oriented, so most results are returned as raster layers. In this step, you will generate the raster for your study area and define the resolution, or pixel size.

For the pixel resolution, the function only accepts values in **decimal degrees**. For example, if you plan to use the results together with WorldClim or CHELSA variables at **30-arc-second** resolution, you should use **0.008333 degrees**. The same logic applies to the other common resolutions: 2.5 arc-minutes = 0.041667, 5 arc-minutes = 0.083333, and 10 arc-minutes = 0.166667. Once you create the vilma.dist object, you can check the results using the function **plot** or **view.vilma** for an interactive view.

<br>
```{r}
library(vilma)

# Create the vilma.dist object, for more information try ?points_to_raster
prim_dist <- points_to_raster(points = occ_final, res = 0.8)
```

```{r, fig.align='center', fig.dpi=230, eval=run_leaflet}
# See the results
view.vilma(prim_dist)

# or use plot(prim_dist)
```
<br>

### 3. Beta-PD indices

Now, you can calculate all beta phylogenetic diversity indices available in `vilma`. Each beta-PD function creates a `vilma.beta` object. This is a large S4 object that contains the pairwise beta-diversity matrix, the list of raster outputs, summary statistics, and information about the algorithm used.

Because beta-diversity indices are calculated as pairwise comparisons among cells, `vilma` provides three different options to transform these values into raster layers. The first option is to calculate the mean beta-diversity value of each cell relative to all other cells. The second option is to perform a Principal Coordinates Analysis (PCoA) and use the first two axes as raster outputs. The third option is to perform a Non-metric Multidimensional Scaling (NMDS) using two axes. Therefore, the statistics associated with these ordination methods, such as PCoA eigenvalues and NMDS stress, are also saved in the `vilma.beta` object.

#### 3.1. Between-Community Mean Pairwise Distance (betaMPD)

**Between-community Mean Pairwise Distance (betaMPD)** is a beta phylogenetic diversity index that measures the average phylogenetic distance between species from two different communities, sites, or raster cells. While alpha MPD summarizes the mean phylogenetic distance among species within a single cell, betaMPD compares species across pairs of cells. Therefore, betaMPD assesses how **phylogenetically different** two communities are.

The logic of betaMPD is based on pairwise comparisons between all species in one cell and all species in another cell. If the species in two cells belong to distant evolutionary lineages, betaMPD will be high. If the species in the two cells are closely related, betaMPD will be low. Thus, larger betaMPD values indicate greater phylogenetic differentiation between communities, whereas smaller values indicate greater phylogenetic similarity.

In `vilma`, betaMPD is calculated with the function `beta.mpd`. This function requires a rooted phylogenetic tree with branch lengths and a `vilma.dist` object containing the species distribution across raster cells. This function takes different arguments that you must consider. 

The function can weight the betaMPD by abundance using the argument `abundance` (default is `FALSE`). Also, the argument `exclude.conspecific` controls whether comparisons involving the same species across two different cells are included. If `exclude.conspecific = FALSE`, conspecific matches are included in the calculation. If `exclude.conspecific = TRUE`, those same-species comparisons are removed, allowing the index to focus only on distances among different species across cells. Like alpha-mpd, this function handles single-species cells with the argument `mpd.method`, using the same three options `exclude`, `root`, and `node`.

The function also includes the argument `normalize`. If `normalize = TRUE`, the between-community betaMPD value is divided by the mean within-cell MPD of the two cells being compared. This produces a relative, dimensionless value that expresses between-cell phylogenetic distance in relation to the internal phylogenetic structure of the two communities. However, this normalized value is **not necessarily restricted to a 0–1 range**. Finally, the argument `scale01` can be used to rescale the final betaMPD matrix between 0 and 1. This option is useful for visualization, but it should be interpreted as a display transformation rather than a change in the biological meaning of the index. Because the scaling depends on the maximum value in the dataset, scaled values **should not be directly compared across different datasets or phylogenies**.

<br>
```{r, fig.align='center', fig.dpi=230, eval=FALSE}

# Calculate the betaMPD
prim_bmpd <- beta.mpd(tree = prim_tree, dist = prim_dist, mpd.method = "root", 
                      abundance = FALSE, exclude.conspecific = FALSE, 
                      normalize = FALSE, scale01 = TRUE)
# Check the summary results
print(prim_bmpd)
```

```{r, fig.align='center', fig.dpi=230, echo=FALSE, warning=FALSE, message=FALSE, eval = run_heavy}

# Calculate the betaMPD
invisible(
  capture.output(
    prim_bmpd <- beta.mpd(tree = prim_tree, dist = prim_dist, mpd.method = "root", 
                      abundance = FALSE, exclude.conspecific = FALSE, normalize = FALSE, scale01 = TRUE)
            )
  )

# Check the summary results
print(prim_bmpd)
```

```{r, fig.align='center', fig.dpi=230, eval=run_leaflet}
# Check the interactive map (see the different available layers)
view.vilma(prim_bmpd)
```
<br>

#### 3.2. Between-community Mean Nearest Taxon Distance (betaMNTD)

**Between-community Mean Nearest Taxon Distance (betaMNTD)** is a beta phylogenetic diversity index that measures the average phylogenetic distance between each species in one community and its nearest phylogenetic relative in another community, site, or raster cell. While alpha MNTD summarizes the mean distance from each species to its closest relative within a single cell, betaMNTD compares nearest relatives across pairs of cells. Therefore, betaMNTD assesses how **phylogenetically different two communities are at the terminal or tip level of the phylogeny**.

The logic of betaMNTD is based on nearest-neighbor comparisons between communities. For each species in one cell, the function identifies the closest relative present in the other cell and calculates that phylogenetic distance. The same procedure is then performed in the opposite direction, from the second cell to the first, and both directions are averaged. If species in two cells have close relatives in each other, betaMNTD will be low. If species in one cell do not have close relatives in the other cell, betaMNTD will be high. Thus, larger betaMNTD values indicate greater phylogenetic differentiation near the tips of the tree, whereas smaller values indicate greater phylogenetic similarity.

In `vilma`, betaMNTD is calculated with the function `beta.mntd`. This function requires a rooted phylogenetic tree with branch lengths and a `vilma.dist` object containing the species distribution across raster cells. This function takes different arguments that you must consider.

The function can weight betaMNTD by abundance using the argument `abundance`; the default is `FALSE`. Also, the argument `exclude.conspecific` controls whether comparisons involving the same species across two different cells are included. If `exclude.conspecific = FALSE`, conspecific matches are included in the calculation. If `exclude.conspecific = TRUE`, those same-species comparisons are removed, allowing the index to focus only on distances among different species across cells. Like alpha MNTD, this function handles single-species cells with the argument `mntd.method`, using the same three options: `exclude`, `root`, and `node`.

The function also includes the argument `normalize`. If `normalize = TRUE`, the between-community betaMNTD value is divided by the mean within-cell MNTD of the two cells being compared. This produces a relative, dimensionless value that expresses between-cell nearest-taxon distance in relation to the internal phylogenetic structure of the two communities. However, this normalized value is **not necessarily restricted to a 0–1 range**. Finally, the argument `scale01` can be used to rescale the final betaMNTD matrix between 0 and 1. This option is useful for visualization, but it should be interpreted as a display transformation rather than a change in the biological meaning of the index. Because the scaling depends on the maximum value in the dataset, scaled values **should not be directly compared across different datasets or phylogenies**.

<br>
```{r, fig.align='center', fig.dpi=230, eval=FALSE}

# Calculate the betaMPD
prim_bmntd <- beta.mntd(tree = prim_tree, dist = prim_dist, mntd.method = "root", 
                        abundance = FALSE, exclude.conspecific = FALSE, 
                        normalize = FALSE, scale01 = TRUE)
# Check the summary results
print(prim_bmntd)
```

```{r, fig.align='center', fig.dpi=230, echo=FALSE, message=FALSE, warning=FALSE, eval = run_heavy}

# Calculate the betaMPD
invisible(
  capture.output(
    prim_bmntd <- beta.mntd(tree = prim_tree, dist = prim_dist, mntd.method = "root", 
                        abundance = FALSE, exclude.conspecific = FALSE, 
                        normalize = FALSE, scale01 = TRUE)
            )
  )

# Check the summary results
print(prim_bmntd)
```

```{r, fig.align='center', fig.dpi=230, eval=run_leaflet}
# Check the interactive map (see the different available layers)
view.vilma(prim_bmntd)
```
<br>

#### 3.3. Phylogenetic Beta Diversity Partitioning (phylobeta)

**Phylogenetic Beta Diversity Partitioning** is a beta-diversity approach that measures the dissimilarity in evolutionary history between two communities, sites, or raster cells. In `vilma`, this index is calculated with the function `phylo.beta`, which partitions total phylogenetic beta diversity into three components: **total dissimilarity**, **turnover**, and **nestedness**.This function requires a rooted phylogenetic tree with branch lengths and a vilma.dist object containing the species distribution across raster cells. The function compares communities using branch-length information from the tree and returns pairwise distance matrices for total dissimilarity, turnover, and nestedness.

The logic of `phylo.beta` is based on comparing the amount of branch length shared and not shared between two communities. **Total phylogenetic dissimilarity**, represented by β_sor, measures the overall difference in evolutionary history between two cells. The **turnover component**, β_sim, represents the replacement of evolutionary lineages between cells. The **nestedness** component, β_sne, represents the portion of the dissimilarity attributable to one community containing a subset of the evolutionary history present in another community. Therefore, `phylo.beta` is useful not only for measuring how different two cells are, but also for understanding whether that difference is mainly due to lineage replacement or to differences in the amount of evolutionary history present in each cell.

The function includes the argument `method`, which can be set to `unweighted` or `weighted`. If `method = "unweighted"`, the calculation is based on species presence or absence on the phylogeny. In this case, the index evaluates whether branches are shared or unique between communities. If `method = "weighted"`, the calculation incorporates abundance information by weighting branches according to the relative abundance of the species they descend from.

The argument normalize is used when `method = "weighted"`. If `normalize = TRUE`, which is the default, within-cell abundances are converted to relative abundances before branch-level values are calculated. This allows the indices to remain within the range from 0 to 1 and makes communities with different total numbers of records more comparable. If `normalize = FALSE`, raw counts are used, which may emphasize differences in sampling intensity or total abundance among cells.

The output of `phylo.beta` is a `vilma.beta` object. This object contains three pairwise distance matrices: total.dissimilarity for β_sor, turnover for β_sim, and nestedness for β_sne. It also includes raster outputs summarizing the mean value of each component per cell, as well as PCoA and NMDS ordination axes mapped into geographic space. The associated statistics, such as PCoA eigenvalues and NMDS stress values, are also saved in the object.


<br>
```{r, fig.align='center', fig.dpi=230, eval=FALSE}

# Calculate the betaMPD
prim_phyb <- phylo.beta(tree = prim_tree, dist = prim_dist, method = "unweighted")
# Check the summary results
print(prim_phyb)
```

```{r, fig.align='center', fig.dpi=230, echo=FALSE, message=FALSE, warning=FALSE, eval = run_heavy}

# Calculate the betaMPD
invisible(
  capture.output(
    prim_phyb <- phylo.beta(tree = prim_tree, dist = prim_dist, method = "unweighted")
            )
  )

# Check the summary results
print(prim_phyb)
```

```{r, fig.align='center', fig.dpi=230, eval=run_leaflet}
# Check the interactive map (see the different available layers)
view.vilma(prim_phyb)
```
<br>

#### 3.4. Phylogenetic Sørensen Similarity (PhyloSor)

**Phylogenetic Sørensen Similarity (PhyloSor)** is a beta phylogenetic diversity index that measures the extent to which evolutionary history is shared between two communities, sites, or raster cells. While traditional Sørensen similarity compares communities based only on shared species, PhyloSor extends this idea by comparing the amount of **shared branch length** in a phylogenetic tree. Therefore, PhyloSor assesses the phylogenetic similarity between two communities.

The logic of PhyloSor is based on the proportion of phylogenetic diversity shared by two communities. If two cells share many branches of the phylogeny, their PhyloSor similarity will be high. If they share little evolutionary history, their similarity will be low. Thus, larger PhyloSor values indicate greater phylogenetic similarity between communities, whereas smaller values indicate greater phylogenetic dissimilarity. The complementary dissimilarity is calculated as 1 - similarity.

In `vilma`, PhyloSor is calculated with the function `phylosor.calc`. This function requires a rooted phylogenetic tree with branch lengths and a `vilma.dist` object containing the species distribution across raster cells. The function returns both a PhyloSor similarity matrix and a complementary dissimilarity matrix.

The function includes the argument `method`, which controls how phylogenetic diversity is calculated for single-species cells. The available options are `root`, `node`, and `exclude`. The argument `singleton_overlap` controls how the function treats pairs of cells that share exactly one species. If `singleton_overlap = FALSE`, that single shared species contributes to shared phylogenetic diversity according to the selected method. If `singleton_overlap = TRUE`, pairs sharing only one species are treated as having no shared phylogenetic diversity, and their similarity is set to zero. This option is important because single-species overlap can strongly influence similarity values in sparse datasets.

The function can also calculate abundance-weighted PhyloSor with the abundance argument. If abundance = FALSE, which is the default, the calculation uses presence-absence data. If abundance = TRUE, branch contributions are weighted by species abundances within each cell. When abundance is used, the argument normalize controls whether abundances are converted to relative abundances before calculating similarity. If normalize = TRUE, the default, each cell is standardized so that its abundances sum to 1, yielding similarity values between 0 and 1. If normalize = FALSE, raw counts are used, which may emphasize differences in sampling intensity or total abundance among cells.

The output of `phylosor.calc` is a vilma.beta object. This object contains the PhyloSor similarity matrix, the complementary dissimilarity matrix, raster layers summarizing the mean similarity and mean dissimilarity for each cell, and ordination rasters from PCoA and NMDS. The associated statistics, such as PCoA eigenvalues and NMDS stress values, are also saved.

<br>
```{r, fig.align='center', fig.dpi=230, eval=FALSE}

# Calculate the betaMPD
prim_phys <- phylosor.calc(tree = prim_tree, dist = prim_dist, singleton_overlap = FALSE, 
                           method = "root", abundance = FALSE, normalize = FALSE)
# Check the summary results
print(prim_phys)
```

```{r, fig.align='center', fig.dpi=230, echo=FALSE, message=FALSE, warning=FALSE, eval = run_heavy}

# Calculate the betaMPD
invisible(
  capture.output(
    prim_phys <- phylosor.calc(tree = prim_tree, dist = prim_dist, singleton_overlap = FALSE, 
                           method = "root", abundance = FALSE, normalize = FALSE)

            )
  )

# Check the summary results
print(prim_phys)
```

```{r, fig.align='center', fig.dpi=230, eval=run_leaflet}
# Check the interactive map (see the different available layers)
view.vilma(prim_phys)
```
<br>

#### 3.5. Rao Beta phylogenetic dissimilarity (Rao beta)

**Rao Beta** phylogenetic dissimilarity is a beta phylogenetic diversity index that measures the expected phylogenetic dissimilarity between two individuals drawn from two different communities, sites, or raster cells. While Rao’s alpha diversity evaluates phylogenetic dissimilarity within a single community, Rao Beta compares phylogenetic composition between pairs of communities. Therefore, Rao Beta assesses how **phylogenetically different two communities are while considering the relative contribution of species within each cell**.

The logic of Rao Beta is based on comparing the phylogenetic distances among species across communities while also accounting for the phylogenetic diversity already present within each community. If two cells contain similar species or closely related lineages, Rao Beta will be low. If two cells contain species from distant evolutionary lineages, Rao Beta will be high. Thus, larger Rao Beta values indicate greater phylogenetic dissimilarity or turnover between communities, whereas smaller values indicate greater phylogenetic similarity.

In `vilma`, Rao Beta is calculated with the function `rao.beta`. This function requires a rooted phylogenetic tree with branch lengths and a `vilma.dist` object containing the species distribution across raster cells. The function can calculate Rao Beta using presence-absence data or abundance information. If `abundance = FALSE`, which is the default, the cell-by-species table is converted to presence-absence before calculating proportions. 

The function also includes the argument `scale01`. If `scale01 = TRUE`, which is the default, the patristic distance matrix is divided by its maximum off-diagonal value, so phylogenetic distances range from 0 to 1. Under this scaling, Rao Beta values also range approximately from 0 to 1. A value close to 0 indicates very similar phylogenetic composition between two cells, while a value close to 1 indicates strong phylogenetic dissimilarity or maximal turnover.

The output of rao.beta is a `vilma.beta` object. This object contains the pairwise Rao Beta dissimilarity matrix, a raster layer summarizing the mean dissimilarity of each cell relative to all other cells, and ordination rasters based on PCoA and NMDS. The associated statistics, such as PCoA eigenvalues and NMDS stress values, are also saved in the object.

<br>
```{r, fig.align='center', fig.dpi=230, eval=FALSE}

# Calculate the betaMPD
prim_raob <- rao.beta(tree = prim_tree, dist = prim_dist, abundance = FALSE, scale01 = TRUE)
# Check the summary results
print(prim_raob)
```

```{r, fig.align='center', fig.dpi=230, echo=FALSE, message=FALSE, warning=FALSE, eval = run_heavy}

# Calculate the betaMPD
invisible(
  capture.output(
    prim_raob <- rao.beta(tree = prim_tree, dist = prim_dist, abundance = FALSE, scale01 = TRUE)
            )
  )

# Check the summary results
print(prim_raob)
```

```{r, fig.align='center', fig.dpi=230, eval=run_leaflet}
# Check the interactive map (see the different available layers)
view.vilma(prim_raob)
```
<br>

#### 3.6. Unique Fraction Metric (UniFrac)

**UniFrac** phylogenetic dissimilarity is a beta phylogenetic diversity index that measures how different two communities, sites, or raster cells are based on the amount of evolutionary history that is unique to each community. Unlike traditional beta-diversity indices that compare only species identities, UniFrac uses the branch lengths of a phylogenetic tree to evaluate whether two communities share or differ in their evolutionary lineages.

The logic of UniFrac is based on the proportion of branch length that is **unique to each community**. If two cells share most of their evolutionary history, their UniFrac distance will be low. If each cell contains lineages not present in the other, their UniFrac distance will be high. Thus, larger UniFrac values indicate greater phylogenetic dissimilarity between communities, whereas smaller values indicate greater phylogenetic similarity.

In `vilma`, UniFrac is calculated with the function `unifrac.calc`. This function requires a rooted phylogenetic tree with branch lengths and a `vilma.dist` object. The function includes the argument `method`, which can be set to `unweighted` or `weighted`. If `method = "unweighted"`, UniFrac is calculated using presence-absence data. In this case, the index evaluates the fraction of the branch length unique to each community relative to the total branch length shared by both communities. This version is useful when the focus is on differences in lineage composition, regardless of abundance. If `method = "weighted"`, UniFrac incorporates relative abundance information. In this case, the function compares the abundance-weighted fraction of each branch across communities. Therefore, weighted UniFrac not only considers whether lineages are present or absent, but also how strongly they are represented within each cell. This version is useful when differences in dominance or relative abundance among lineages are important.

In both cases, UniFrac values range from 0 to 1. Values close to 0 indicate that two cells share most of their phylogenetic composition, while values close to 1 indicate strong phylogenetic dissimilarity or high lineage turnover. The unweighted version emphasizes differences in the presence or absence of evolutionary lineages, while the weighted version emphasizes differences in their relative representation.

The output of `unifrac.calc` is a `vilma.beta` object. This object contains the pairwise UniFrac dissimilarity matrix, a raster layer summarizing the mean UniFrac dissimilarity of each cell relative to all other cells, and ordination rasters based on PCoA and NMDS. The associated statistics, such as PCoA eigenvalues and NMDS stress values, are also saved in the object.

<br>
```{r, fig.align='center', fig.dpi=230, eval=FALSE}

# Calculate the betaMPD
prim_uni <- unifrac.calc(tree = prim_tree, dist = prim_dist, method = "weighted")
# Check the summary results
print(prim_uni)
```

```{r, fig.align='center', fig.dpi=230, echo=FALSE, message=FALSE, warning=FALSE, eval = run_heavy}

# Calculate the betaMPD
invisible(
  capture.output(
    prim_uni <- unifrac.calc(tree = prim_tree, dist = prim_dist, method = "weighted")
            )
  )

# Check the summary results
print(prim_uni)
```

```{r, fig.align='center', fig.dpi=230, eval=run_leaflet}
# Check the interactive map (see the different available layers)
view.vilma(prim_uni)
```
<br>

### 4. Saving results.

Once you have selected the index or indices that you want to keep and use for further analyses, `vilma` allows you to save the beta phylogenetic diversity results using the function `write.vilma.beta`. This function requires a `vilma.beta` object, a `file` name, which can also include the output directory path, and the `raster.format` argument. The available raster format options are `tif`, `grd`, and 
`.

The function generates different output files. First, it creates a **CSV file** containing the values associated with each raster cell, including species richness, phylogenetic diversity values, and any other index-specific results. Second, it saves the **raster layers** available in the `vilma.beta` object using the raster format specified by the user. Third, it generates a **log file** containing the summary information of the `vilma.beta` object, equivalent to the information displayed when using the print function. The numbe of files will depend on the index used, please see help panel for further information (`?write.vilma.beta`) Let see this example saving the results of `rao.beta``.

<br>
```{r, eval=FALSE}

# Saving the results as
write.vilma.pd(vilma.beta = prim_raob, file = "results/Prim_Madagascar", 
               raster.format = "tif", overwrite = TRUE)

```
<br>

The output folder will looks like:

<br>
```{}
results/
  ├── Prim_Madagascar_RaoBeta.csv
  ├── Prim_Madagascar__mean.dissimilarity.tif
  ├── Prim_Madagascar__ndms.1.tif
  ├── Prim_Madagascar__ndms.2.tif
  ├── Prim_Madagascar__pcoa.1.tif
  ├── Prim_Madagascar__pcoa.2.tif
  └── Prim_Madagascar_raoB_log.txt
```
<br>

---

### 5. Conclusion

In this tutorial, you learned how to calculate beta phylogenetic diversity indices in `vilma` using occurrence data, a phylogenetic tree, and a `vilma.dist` object. You explored several indices, including betaMPD, betaMNTD, phylogenetic beta-diversity partitioning, PhyloSor, Rao Beta, and UniFrac.

Each index captures a different aspect of phylogenetic variation among raster cells, such as lineage turnover, shared evolutionary history, nearest-taxon differences, nestedness, or abundance-weighted dissimilarity. Because beta diversity is calculated through pairwise comparisons among cells, vilma summarizes the results as distance or similarity matrices, mean beta-diversity rasters, and ordination layers based on PCoA and NMDS.

Finally, you learned how to save beta-diversity results using `write.vilma.beta`, including raster outputs, cell-level tables, and log files. Overall, `vilma` provides a flexible raster-based framework to evaluate how evolutionary history changes across space and to identify areas with distinct phylogenetic composition.

